---
title: "Stat 850 Project Description"
author: "David Nguyen"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
options(digits = 3)
```

```{r load-sas-libraries, echo = F, message = F, warning = F}

saspath <- "/usr/local/SASHome/SASFoundation/9.4/bin/sas_en" # Your SAS path goes here
sasopts <- "-ls 75"
knitr::opts_chunk$set(engine.path=list(sas=saspath, saslog=saspath, sashtml=saspath),
                      engine.opts=list(sas=sasopts, saslog=sasopts, sashtml=sasopts), 
                      comment=NA)
library(SASmarkdown)
```

## Instructions

Each member of your team should modify this document in some way and push their modifications to the repository in a separate commit. This will ensure that you have set your repository up in a way that ensures all group members are working with the same repository. 

Note that you must compile your readme (this document) for it to properly display as part of your github repository. 

Once you have received feedback on your project proposal (via Canvas) you may alter this README so that it describes your final project instead of the project proposal. 

## Data Set
Original pdf of [postal data](https://www.courtlistener.com/recap/gov.uscourts.nysd.542242/gov.uscourts.nysd.542242.45.1.pdf).
These data were cited by [the Guardian](https://www.theguardian.com/us-news/2020/sep/21/usps-post-office-mail-slowdowns-louis-dejoy) as evidence that 

> "The United States Postal Service (USPS) saw a severe decline in the rate of on-time delivery of first-class mail after Louis DeJoy took over as postmaster general."

I've uploaded the cleaned and formatted version of the data as mail_on_time.csv. The data set contains 8325 rows and 5 columns.

The variables are:

* Date: self-explanatory
* Area: The region of interest ("Nation", "Capital Metro", "Eastern", "Great Lakes", "Northeast", 
"Pacific", "Southern", "Western")
* District: mail district within an Area (e.g., "Atlanta", "Baltimore", "Capital", "Greater S Carolina", 
"Greensboro", "Mid-Carolinas" ...)
* mail_type: If the mail is first-class, marketing, or periodical
* pct_on_time: The percent of mail delivered on time

## Potential Topics to Explore Using the Data Set
Describe avenues you might explore using the data

I'm interested in looking at USPS performance data (pct_on_time) over the past year to detect if there changes that may have been caused by the SARS-CoV-2 pandemic or by Louis DeJoy's operational changes.

I am interested in looking at the timing of changes in USPS performance with respect to COVID-19 shutdown and the appointment of Louis DeJoy as well as spatial variation in changes in USPS performance. I hope to create interactive line plots and maps.

## Group Members
David Nguyen

